Calibrating seed-based alignment heuristics with Sesame

### 2020-02-27 manuscript.tex
Add explanations for skip seeds

\begin{abstract}
% Statement of the problem.
The increasing throughput of DNA sequencing technologies creates a need
for faster algorithms. The fate of most reads is to be mapped to a
reference sequence, typically a genome. Modern mappers rely on heuristics
to gain speed at a reasonable cost for accuracy. In the seeding heuristic,
short matches between the reads and the genome are used to narrow the
search to a set of candidate locations. Several seeding variants used in
modern mappers show good empirical performance but they are difficult to
calibrate or to optimize for lack of theoretical results.
% Summary of the results.
Here we develop a theory to estimate the probability that the correct
location of a read is filtered out during seeding, resulting in mapping
errors. We describe the properties of simple exact seeds, skip-seeds and
MEM seeds (Maximal Exact Match seeds).
% Innovation.
The main innovation of this work is to use concepts from analytic
combinatorics to represent reads as abstract sequences, and to specify
their generative function to estimate the probabilities of interest.
% Durability.
We provide several algorithms, which together give a workable solution for
the problem of calibrating seeding heuristics for short reads. We also
provide a C implementation of these algorithms in a library called Sesame.
% Scope.
These results can improve current mapping algorithms and lay the
foundation of a general strategy to tackle sequence alignment problems.
% Availability
The Sesame library is open source and available for download at
\url{https://github.com/gui11aume/sesame}.
\end{abstract}

@@

\begin{definition}
\label{def:seg}
A terminator is any symbol that is different from the symbol $\square$. A
segment is a sequence of 0 or more symbols $\square$ followed by a
terminator. The tail is the last segment of the read, where the terminator
is always the special symbol $|$.
\end{definition}

@@

The probability of occurrence of the symbol $\Downarrow$ is $p$ (the error
rate of the sequencer) so the probability of occurrence of the symbol
$\square$ is $1-p = q$. Both symbols have size 1, so their respective
weighted generating function are $pz$ and $qz$. Using the rule for
concatenation, we see that a segment of $i$ symbols $\square$ followed by
a terminator has weighted generating function $(qz)^ipz$. The final symbol
$|$ has size 0, so a tail segment of $i$ symbols $\square$ followed by the
symbol $|$ has weighted generating function $(qz)^i$.

@@

In the representation above, the different types of segments are
identified by their terminator. The terminators are written in the margins
of the transfer matrix for readability.

@@

As in the previous example, we need to know whether a nucleotide is a
sequencing error, but this time we also need to know its phase in the
repeated cycles of skipped positions. For this, we define the skip-$n$
alphabet $\mathcal{A}_n = \{\square, |, \Downarrow_0, \Downarrow_1,
\ldots, \Downarrow_n \}$. Again, the symbol $\square$ represents a correct
nucleotide and the symbol $|$ is a terminator added at the end of the
read. The symbols $\Downarrow_i$ ($i \in \mathbb{N}$) represent sequencing
errors and $i$ indicates the number of nucleotides till the next
non-skipped position (\textit{i.e.} $i=0$ for nucleotides immediately
before a non-skipped position and $i=n$ for nucleotides at a non-skipped
position).

As per definition~\ref{def:seg}, segments in this alphabet are sequences
of 0 or more symbols $\square$ followed by any of the symbols
$\Downarrow_i$ or by the symbol $|$. Given that this decomposition is
unique, we can again view a read as a sequence of segments with a tail.
The example of Figure~\ref{fig:simple} is shown again in
Figure~\ref{fig:skip}, where segments in the mismatch alphabet have been
replaced by segments in the skip-3 alphabet.

\begin{figure}[h]
\centering
\includegraphics[scale=0.85]{sketch_skip.pdf}
\caption{\textbf{The skip encoding.}
The read of Figure~\ref{fig:simple} is represented in the skip-3 alphabet.
The symbols $\Downarrow_i (i = 1,2,3)$ represent mismatches against the
target (they are erroneous nucleotides) and the symbol $\square$
represents a match (it is a correct nucleotide). The vertical bars
indicate non-skipped positions (the potential start of a seed). The number
$i$ in $\Downarrow_i$ indicates the number of nucleotides till the next
non-skipped position. For $i=0$ the next nucleotide is not skipped.
Other features are as in Figure~\ref{fig:simple}.}
\label{fig:skip}
\end{figure}

The probability of occurrence of a sequencing error is $p$, so every
symbol $\Downarrow_i$ has the same weighted generating function $pz$ ---
implicitly assuming that the next non-skipped position is at distance $i$,
otherwise the weighted generating function is 0. The weighted generating
function of the symbol $\square$ is again $qz$, and so the weighted
generating function of a segment with $i$ symbols $\square$ is $(qz)^ipz$.
Likewise, the weighted generating function of a tail with $i$ symbols
$\square$ is $(qz)^i$.

Reads that do not contain any on-target skip seed can contain segments
with $\gamma$ or more symbols $\square$, so this case is more complex than
the previous one. For instance, if there is a sequencing error $i$
nucleotides before the next non-skipped position, the status of the next
$i$ nucleotides does not have any influence on the presence of an
on-target seed. Indeed, there is no on-target seed upstream of the error
(by construction) and the next seed is scheduled to start $i$ nucleotides
downstream, so there can be up to $\gamma+i-1$ symbols $\square$ in a row.
In order to enforce the absence of on-target seeds, we thus have to adjust
the maximum size of the segments depending on the preceding terminator.

In the skip-$n$ alphabet, reads are composed of segments terminated by the
$n+1$ symbols $\Downarrow_0, \Downarrow_1, \ldots, \Downarrow_n$, plus the
tail. The dimensions of the transfer matrix are thus $(n+2) \times (n+2)$.

As mentioned above, after the terminator $\Downarrow_i$, we can append
segments with up to $\gamma+i-1$ symbols $\square$. The terminators of
those $\gamma+i$ possible segments are distributed according to the laws
of the arithmetic modulo $n+1$. When $i > 0$, for instance, if the next
segment has no symbol $\square$ then it is terminated by
$\Downarrow_{i-1}$. But the same goes if the segment has $n+1$ symbols
$\square$ (assuming $n+1 < \gamma$). Specifying the transfer matrix thus
involves a substantial amount of bookkeeping. One can check that the
expression of the transfer matrix $M_n(z)$ is
\begin{equation*}
\begin{blockarray}{cccccc}
   & \Dn{0} & \Dn{1} & \ldots & \Dn{n} & \nd \\
\begin{block}{c[ccccc]}
\Dn{0} & H_{0,0}(z) & H_{0,1}(z) & \ldots & H_{0,n}(z) & J_0(z) \\
\Dn{1} & H_{1,0}(z) & H_{1,1}(z) & \ldots & H_{1,n}(z) & J_1(z) \\
\vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
\Dn{n} & H_{n,0}(z) & H_{n,1}(z) & \ldots & H_{n,n}(z) & J_n(z) \\
\nd & 0 & 0 & \ldots & 0 & 0 \\
\end{block}
\end{blockarray},
\end{equation*}
\begin{eqnarray}
H_{i,j}(z) &=& pz \cdot (qz)^x \cdot \big( 1 + (qz)^{n+1} +
  \ldots + (qz)^{m(n+1)} \big), \\
\notag
  &\;& \text{where } x = i-j-1 \modulo{n+1},
  \text{ and } m = \left\lfloor
  \frac{\gamma+i-1-x}{n+1} \right\rfloor, \\
J_i(z) &=& 1 + qz + (qz)^2 + \ldots + (qz)^{\gamma+i-1},
\end{eqnarray}
where $\lfloor \ldots \rfloor$ denotes the ``floor'' function.

The weighted generating function of interest is the top right entry of the
matrix $M_n(z) + M_n(z)^2 + \ldots = M_n(z)\cdot(I-M_n(z))^{-1}$. To see
why, observe that, at the start of every read, the next nucleotide is a
non-skipped position, so every read can be prepended by $\Downarrow_0$
segments and only by those. Thus, reads are precisely the sequences of
segments that can follow the symbol $\Downarrow_0$ and that are terminated
by a tail, whose weighted generating function is the entry of the matrix
associated with terminators $\Downarrow_0$ and $|$.

$M_n(z)$ is too complex to allow a closed expression of $M_n(z) \cdot
(I-M_n(z))^{-1}$ to be computed. We return to the definition $M_n(z) +
M_n(z)^2 + \ldots$ and observe that the terms $M_n(z)^{k+2}, M_n(z)^{k+3},
\ldots$ have no influence on the coefficients of the weighted generating
function $a_0, a_1, \ldots, a_k$.

Indeed, a read of size $k$ has at most $k+1$ segments (including the
obligatory tail). Since $M(z)^{s+1}$ contains the weighted generating
functions of reads with exactly $s+1$ segments including the tail, $a_k$
cannot depend on $M(z)^{k+2}, M(z)^{k+3}, \ldots$. More formally, we can
prove by induction that all the entries of $M(z)^k$ are divisible by
$z^{k-1}$, showing that the contribution of $M(z)^{k+2} + M(z)^{k+3} +
\ldots$ to $a_0 + a_1z + \ldots +a_kz^k$ is strictly $0$.

So we can compute the matrix $M_n(z) + M_n(z)^2 + \ldots + M_n(z)^{k+1}$,
extract the entry of interest and then compute the terms of the Taylor
expansion up to order $k$. But we can do better: since we are only
interested in the coefficients up to order $k$, we can perform all
algebraic operations on truncated polynomials of order $k$, \textit{i.e.}
we discard the coefficients of order $k+1$ or greater when multiplying two
polynomials.

@@

For exact and skip-seeds, the conclusions are different. The situation is
even the opposite if we consider reads with a single sequencing error. In
this case the target is always in the candidate set (if the read size is
greater than $2\gamma$, where $\gamma$ is the minimum seed length), so
mapping errors are never due to failures of the seeding heuristic.
Instead, mapping errors happen only when the true location is not the
best.

@@

In conclusion, the true location problem can be calibrated using the
methods presented here when using MEM seeds, and by computing the
probability that the true location is not the best when using exact or
skip-seeds.

@@

>>>>

\begin{abstract}
% Statement of the problem.
The increasing throughput of DNA sequencing technologies creates a need
for faster algorithms. The fate of most reads is to be mapped to a
reference sequence, typically a genome. Modern mappers rely on heuristics
to gain speed at a reasonable cost for accuracy. In the seeding heuristic,
short matches between the reads and the genome are used to narrow the
search to a set of candidate locations. Several seeding variants used in
modern mappers show good empirical performance but they are difficult to
calibrate or to optimize for lack of theoretical results.
% Summary of the results.
Here we develop a theory to estimate the probability that the correct
location of a read is filtered out during seeding, resulting in mapping
errors. We describe the properties of simple exact seeds, skip seeds and
MEM seeds (Maximal Exact Match seeds).
% Innovation.
The main innovation of this work is to use concepts from analytic
combinatorics to represent reads as abstract sequences, and to specify
their generative function to estimate the probabilities of interest.
% Durability.
We provide several algorithms, which together give a workable solution for
the problem of calibrating seeding heuristics for short reads. We also
provide a C implementation of these algorithms in a library called Sesame.
% Scope.
These results can improve current mapping algorithms and lay the
foundation of a general strategy to tackle sequence alignment problems.
% Availability
The Sesame library is open source and available for download at
\url{https://github.com/gui11aume/sesame}.
\end{abstract}

@@

\begin{definition}
\label{def:seg}
A terminator is any symbol that is different from the symbol $\square$. A
segment is a sequence of 0 or more $\square$ symbols followed by a
terminator. The tail is the last segment of the read, where the terminator
is always the special symbol $|$.
\end{definition}

@@

The probability of occurrence of the symbol $\Downarrow$ is $p$ (the error
rate of the sequencer) so the probability of occurrence of the $\square$
symbol is $1-p = q$. Both symbols have size 1, so their respective
weighted generating function are $pz$ and $qz$. Using the rule for
concatenation, we see that a segment of $i$ symbols $\square$ followed by
a terminator has weighted generating function $(qz)^ipz$. The final symbol
$|$ has size 0, so a tail segment of $i$ symbols $\square$ followed by the
symbol $|$ has weighted generating function $(qz)^i$.

@@

In the representation above, the different types of segments are
identified by their terminator, indicated in the margins for clarity.

@@

We need to know whether a nucleotide is a sequencing error, but this time
we also need to know its phase in the repeated cycles of skipped
positions. For this, we define the skip-$n$ alphabet $\mathcal{A}_n =
\{\square, |, \Downarrow_0, \Downarrow_1, \ldots, \Downarrow_n \}$. Again,
the symbol $\square$ represents a correct nucleotide and the symbol $|$ is
a terminator added at the end of the read. The symbols $\Downarrow_j$ ($j
\in \mathbb{N}$) represent sequencing errors and $j$ indicates the number
of nucleotides till the next non-skipped position (\textit{i.e.} $j=0$ for
nucleotides immediately before a non-skipped position and $j=n$ for
nucleotides at a non-skipped position).

As per definition~\ref{def:seg}, segments in this alphabet are sequences
of 0 or more $\square$ symbols followed by any of the symbols
$\Downarrow_j$ or by the symbol $|$. Given that this decomposition is
unique, we can again view a read as a sequence of segments with a tail.
The example of Figure~\ref{fig:simple} is shown again in
Figure~\ref{fig:skip}, where segments in the mismatch alphabet have been
replaced by segments in the skip-3 alphabet.

\begin{figure}[h]
\centering
\includegraphics[scale=0.85]{sketch_skip.pdf}
\caption{\textbf{The skip encoding.}
The read of Figure~\ref{fig:simple} is represented in the skip-3 alphabet.
The symbols $\Downarrow_j (j = 1,2,3)$ represent mismatches against the
target (they are erroneous nucleotides) and the $\square$ symbol
represents a match (it is a correct nucleotide). The vertical bars
indicate non-skipped positions (the potential start of a seed). The number
$j$ in $\Downarrow_j$ indicates the number of nucleotides till the next
non-skipped position. For $j=0$ the next nucleotide is not skipped.
Other features are as in Figure~\ref{fig:simple}.}
\label{fig:skip}
\end{figure}

The probability of occurrence of a sequencing error is $p$, so every
symbol $\Downarrow_j$ has the same weighted generating function
$pz$---provided the next non-skipped position is at distance $j$,
otherwise the weighted generating function is 0. The weighted generating
function of the symbol $\square$ is again $qz$, so a segment with $i$
symbols $\square$ followed by the symbol $\Downarrow_j$ has weighted
generating function $(qz)^ipz$---if the next non-skipped position is at
distance $j$, otherwise it is 0. As in the previous example, a tail
segment with $i$ symbols $\square$ followed by the symbol $|$ has weighted
generating function $(qz)^i$.

This case is more complex than the previous one: reads without on-target
skip seed of minimum size $\gamma$ can contain segments with $\gamma$ or
more $\square$ symbols. For instance, the read shown in
Figure~\ref{fig:skip} contains a stretch of 9 nucleotides without errors
but it has no seed of minimum size $\gamma=9$. More generally, if there is
a sequencing error $i$ nucleotides before the next non-skipped position,
there can be up to $\gamma+i-1$ symbols $\square$ in a row.

Expressed in different words, it is possible to append segments with up to
$\gamma+i-1$ symbols $\square$ after segments terminated by $\Downarrow_i$
($\Downarrow_i$-segments for short). Each of those $\gamma+i$ possible
segments is associated with a different terminator, depending on how far
ahead the next non-skipped position lies. In Figure~\ref{fig:skip}, for
instance, the second segment is terminated by $\Downarrow_1$ because there
is 1 nucleotide before the next non-skipped position. If the segment were
1 nucleotide longer, the terminator would have to be $\Downarrow_0$.

The main issue is that segments of different lengths can be terminated by
the same symbol. Going back to Figure~\ref{fig:skip}, the third segment
has length 10 and is terminated by $\Downarrow_3$. It would also be
terminated by $\Downarrow_3$ if it had length 2 or length 6. In the
general case, the entries of the transfer matrix show some periodicity
modulo $n+1$.

Denote $H_{i,j}(z)$ the weighted generating function of
$\Downarrow_j$-segments that can follow a $\Downarrow_i$-segment. The
total number of segments that can follow a $\Downarrow_i$-segment is
$\gamma+i$. Among them, the shortest $\Downarrow_j$-segment has size
$\ell_0$ to be determined below, and the others have sizes $\ell_0+(n+1),
\ell_0+2(n+1), \ldots, \ell_0+m(n+1)$, for some integer $m$. This $m$ is
the largest number such that $\ell_0+m(n+1) \leq \gamma+i$, so $m =
\lfloor (\gamma+i-\ell_0)/(n+1)\rfloor$, where $\lfloor \ldots \rfloor$ is
the ``floor'' function.

Those segments follow a $\Downarrow_i$-segment, so they start $i$
nucleotides in front of the next non-skipped position. The shortest
following segment that ends $j$ nucleotides before a non-skipped position
has length $\ell_0 = i-j$ if $i > j$, and $n+1-j+i$ otherwise. This is
equivalent to defining $x$ as the number of $\square$ symbols of the
shortest $\Downarrow_j$-segment, with $x = i-j-1 \modulo{n+1}$ and $\ell_0
= x+1$.

The shortest $\Downarrow_j$-segment consists of $x$ symbols $\square$
followed by the $\Downarrow_j$ terminator, so it has weighted generating
function $(qz)^xpz$. The other segments have a multiple of $n+1$
additional $\square$ symbols so their weighted generating functions are
$(qz)^{x+(n+1)}pz, (qz)^{x+2(n+1)}pz, \ldots, (qz)^{x+m(n+1)}pz$. Summing
over those cases, we finally obtain
\begin{eqnarray}
H_{i,j}(z) &=& (qz)^xpz \cdot \big( 1 + (qz)^{n+1} +
  \ldots + (qz)^{m(n+1)} \big), \\
\notag
  &\;& \text{where } x = i-j-1 \modulo{n+1},
  \text{ and } m = \left\lfloor
  \frac{\gamma+i-1-x}{n+1} \right\rfloor.
\end{eqnarray}

Denote $J_i(z)$ the weighted generating function of tails that can follow
a $\Downarrow_i$-segment. There are $\gamma+i$ of them, each consisting of
0 to $\gamma+i-1$ symbols $\square$ followed by the special $|$
terminator for the end of the read. The $|$ has size 0 so its weighted
generating function is 1. Once again summing over the different cases, we
obtain
\begin{eqnarray}
J_i(z) &=& 1 + qz + (qz)^2 + \ldots + (qz)^{\gamma+i-1}.
\end{eqnarray}

Finally, the expression of $M_n(z)$ the transfer matrix of reads in the
skip-$n$ alphabet is
\begin{equation*}
\begin{blockarray}{cccccc}
   & \Dn{0} & \Dn{1} & \ldots & \Dn{n} & \nd \\
\begin{block}{c[ccccc]}
\Dn{0} & H_{0,0}(z) & H_{0,1}(z) & \ldots & H_{0,n}(z) & J_0(z) \\
\Dn{1} & H_{1,0}(z) & H_{1,1}(z) & \ldots & H_{1,n}(z) & J_1(z) \\
\vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
\Dn{n} & H_{n,0}(z) & H_{n,1}(z) & \ldots & H_{n,n}(z) & J_n(z) \\
\nd & 0 & 0 & \ldots & 0 & 0 \\
\end{block}
\end{blockarray}.
\end{equation*}

The weighted generating function of interest is the top right entry of the
matrix $M_n(z) + M_n(z)^2 + \ldots = M_n(z)\cdot(I-M_n(z))^{-1}$. To see
why, observe that, at the start of every read, the next nucleotide is a
non-skipped position, so every read can be prepended by
$\Downarrow_0$-segments and only by those. Thus, reads are precisely the
sequences of segments that can follow the symbol $\Downarrow_0$ and that
are terminated by a tail, whose weighted generating function is the entry
of the matrix associated with terminators $\Downarrow_0$ and $|$.

By construction, the Taylor expansion of the top right term of the matrix
$M_n(z)\cdot(I-M_n(z))^{-1}$ contains the the probabilities of interest.
More specifically, $a_k$, the coefficient of $z^k$ in the series is the
probability that a read of size $k$ contains no skip-$n$ seed of minimum
size $\gamma$.

But $M_n(z)$ is too complex to find a closed expression for $M_n(z) \cdot
(I-M_n(z))^{-1}$. Instead, we return to the definition $M_n(z) + M_n(z)^2
+ \ldots$ and show in appendix~\ref{app_skip} that the coefficient of
interest, $a_k$, only depends on the first $k+1$ terms of the sum. So for
reads of size $k$ or lower, we only need to compute the matrix $M_n(z) +
M_n(z)^2 + \ldots + M_n(z)^{k+1}$ and work out the Taylor expansion of the
top right term.

But we can do better: since we are only interested in the coefficients up
to order $k$, we can perform all algebraic operations on truncated
polynomials of order $k$, \textit{i.e.} we discard the coefficients of
order $k+1$ or greater when multiplying two polynomials.

@@

For exact and skip seeds, the conclusions are different. The situation is
even the opposite if we consider reads with a single sequencing error. In
this case the target is always in the candidate set (if the read size is
greater than $2\gamma$, where $\gamma$ is the minimum seed length), so
mapping errors are never due to failures of the seeding heuristic.
Instead, mapping errors happen only when the true location is not the
best.

@@

In conclusion, the true location problem can be calibrated using the
methods presented here when using MEM seeds, and by computing the
probability that the true location is not the best when using exact or
skip seeds.

@@

\subsection{Derivation of something important}
\label{app_skip}

Easy.

observe that the terms $M_n(z)^{k+2}, M_n(z)^{k+3},
\ldots$ have no influence on the coefficients of the weighted generating
function $a_0, a_1, \ldots, a_k$.

Indeed, a read of size $k$ has at most $k+1$ segments (including the
obligatory tail, which can have size 0). Since $M(z)^{s+1}$ contains the
weighted generating functions of reads with exactly $s+1$ segments
including the tail, $a_k$ cannot depend on $M(z)^{k+2}, M(z)^{k+3},
\ldots$ More formally, we can prove by induction that all the entries of
$M(z)^k$ are divisible by $z^{k-1}$, showing that the contribution of
$M(z)^{k+2} + M(z)^{k+3} + \ldots$ to $a_0 + a_1z + \ldots +a_kz^k$ is
strictly $0$.
